---
title: "Getting Started with epiworldRcalibrate"
subtitle: "Installation and Setup Guide"
author: "Sima Najafzadehkhoei"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-tools: true
    code-copy: true
    highlight-style: github
    df-print: kable
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

## Overview

**epiworldRcalibrate** provides fast, data-driven calibration of SIR epidemic
parameters using a pretrained Bidirectional LSTM (BiLSTM) neural network.
Given a 61-day incidence time series, the package estimates:

- `ptran` — transmission probability per contact
- `crate` — contact rate
- `R0` — basic reproduction number

The package is fully integrated with
[epiworldR](https://github.com/UofUEpiBio/epiworldR) and requires **no
manual Python setup** — everything is handled automatically through
`reticulate`.

::: {.callout-note}
## Why Python?

The calibration model is a PyTorch BiLSTM neural network trained in Python.
R cannot run PyTorch natively, so the package uses
[`reticulate`](https://rstudio.github.io/reticulate/) to delegate inference
to a managed Python environment. You don't need to install Python yourself —
`reticulate >= 1.41` uses `uv` to create and manage a dedicated virtual
environment automatically.
:::

---

## Prerequisites

Before installing `epiworldRcalibrate`, make sure you have the following:

| Requirement | Version | Notes |
|---|---|---|
| R | ≥ 4.1 | [Download R](https://cran.r-project.org/) |
| `devtools` | any | For GitHub installation |
| `reticulate` | **≥ 1.41** | Must support `py_require()` |
| `epiworldR` | any | For running SIR simulations |

Install or update the required R packages first:

```{r}
#| label: prerequisites

# Install devtools if needed
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")

# reticulate >= 1.41 is REQUIRED — older versions lack py_require()
install.packages("reticulate")

# epiworldR for running SIR simulations
install.packages("epiworldR")
```

---

## Installation

Install `epiworldRcalibrate` directly from GitHub:

```{r}
#| label: install

devtools::install_github("sima-njf/epiworldRcalibrate")
```

---

## Python Setup

### Why this step is needed

When you load `epiworldRcalibrate`, it checks whether Python dependencies have
been configured. If they haven't, you will see this startup message:

```
epiworldRcalibrate: Python dependencies are not set up yet.
Run this once to install them:

  epiworldRcalibrate::setup_python_deps()
```

The `setup_python_deps()` function configures a managed Python environment with
the required packages:

- `numpy` — array operations
- `scikit-learn` — data scaling
- `joblib` — model serialization
- `torch` — the PyTorch deep learning backend

This step is intentionally **not automatic** — silently installing Python
packages during normal package use would be intrusive and could break other
Python environments on your system.

::: {.callout-important}
## Run this step once per machine

`setup_python_deps()` only needs to be called **once** after first installing
the package. In subsequent R sessions, Python will be initialized automatically.
:::

### Choosing the right argument

Use `force = TRUE` the **first time** you set up on a new machine. Use
`force = FALSE` (the default) in all subsequent sessions.

::: {.panel-tabset}

## First-time setup

Run this **once**, on a fresh R session, the very first time you install the
package on a machine:

```{r}
#| label: setup-first-time

library(epiworldRcalibrate)

# First time ever: Python is not yet configured — use force = TRUE
setup_python_deps(force = TRUE)
```

**Why `force = TRUE`?** This resets Python requirements from scratch, which is
needed when no managed environment exists yet. It declares all required
packages to `reticulate::py_require()`, which then resolves and installs them
into a dedicated virtual environment backed by `uv`.

::: {.callout-warning}
`force = TRUE` requires that Python has **not yet been initialized** in the
current R session. If Python was already loaded (e.g., you ran some other
`reticulate` code first), you'll see an error asking you to restart R. Just
restart R and run `setup_python_deps(force = TRUE)` again.
:::

## Returning users

In any subsequent R session (after the first-time setup), use `force = FALSE`
or simply omit the argument:

```{r}
#| label: setup-returning

library(epiworldRcalibrate)

# All future sessions: verify packages without resetting
setup_python_deps(force = FALSE)
# or equivalently:
setup_python_deps()
```

**Why `force = FALSE`?** This verifies that all required packages are importable
without touching the existing environment. It's a lightweight check that
confirms your setup is still intact.

:::

### Expected output

When setup completes successfully, you will see:

```
Configuring Python requirements via reticulate::py_require() (uv-backed)...
Verifying installation...
[OK]   numpy v1.26.4
[OK]   sklearn v1.4.2
[OK]   joblib v1.4.0
[OK]   torch v2.3.0

Python setup complete! You can now use the package normally.
Example:
  epiworldRcalibrate::init_bilstm_model()
```

---

## Loading the Model

After Python is set up, load the pretrained BiLSTM model into memory. Do this
once per R session, before running any calibrations:

```{r}
#| label: load-model

init_bilstm_model()
```

```
BiLSTM model loaded successfully.
```

The model and its scalers are bundled inside the package — no external
downloads are needed. You can verify the model status at any time:

```{r}
#| label: check-status

check_model_status()
```

---

## Running a Calibration

### Simulate data with epiworldR

```{r}
#| label: simulate
#| eval: false

library(epiworldR)
library(epiworldRcalibrate)

# Simulate an SIR model
m <- ModelSIRCONN(
  name              = "sim",
  n                 = 8000,
  prevalence        = 0.01,
  contact_rate      = 3,
  transmission_rate = 0.25,
  recovery_rate     = 0.1
)

run(m, ndays = 60)

# Extract the 61-day incidence time series (days 0–60)
inc <- plot_incidence(m)[, 1]
```

### Calibrate SIR parameters

Pass the incidence vector, population size, and recovery rate to
`calibrate_sir()`:

```{r}
#| label: calibrate
#| eval: false

result <- calibrate_sir(
  daily_cases     = inc,
  population_size = 8000,
  recovery_rate   = 0.1
)

print(result)
```

```
    ptran     crate        R0
0.2489123 3.0214567 0.7554231
```

::: {.callout-note}
## Input requirements

- `daily_cases` must be a numeric vector of **exactly 61 values** (days 0
  through 60), all non-negative.
- `population_size` must be a single positive number.
- `recovery_rate` must be a single positive number.
:::

### One-step convenience wrapper

`calibrate_sir()` automatically loads the model if it hasn't been loaded yet
(`auto_init = TRUE` by default), so you can skip `init_bilstm_model()` if you
prefer a single-call workflow:

```{r}
#| label: one-step
#| eval: false

result <- calibrate_sir(
  daily_cases     = inc,
  population_size = 8000,
  recovery_rate   = 0.1
)
```

---

## Diagnostics and Utilities

### Check Python setup

```{r}
#| label: check-python

check_python_setup()
```

Reports Python version, path, and the status of each required package.

### Check model status

```{r}
#| label: check-model

check_model_status()
```

Returns a list with `loaded`, `model_directory`, and `python_config`.

### Free memory

```{r}
#| label: cleanup

cleanup_model()
```

Unloads the model from memory. Useful in long-running sessions or when
switching between model directories.

---

## Quick Reference

| Function | When to call |
|---|---|
| `setup_python_deps(force = TRUE)` | **First time only** — fresh machine, first install |
| `setup_python_deps(force = FALSE)` | Any subsequent session, to verify setup |
| `init_bilstm_model()` | Once per R session, before predicting |
| `calibrate_sir(...)` | Each time you want parameter estimates |
| `check_python_setup()` | Diagnose Python or package issues |
| `check_model_status()` | Check if model is loaded and where |
| `cleanup_model()` | Free memory when done |

---

## Troubleshooting

### "reticulate >= 1.41 required"

```
Error: epiworldRcalibrate requires reticulate >= 1.41 with py_require() support.
Please update reticulate and restart R.
```

**Fix:** Update reticulate and restart R:

```{r}
install.packages("reticulate")
```

### "force = TRUE requires a fresh R session"

```
Error: `force = TRUE` requires a fresh R session before Python initializes.
Please restart R and rerun setup_python_deps(force = TRUE).
```

**Fix:** Restart R, then call `setup_python_deps(force = TRUE)` before running
any other code.

### "daily_cases must have length 61"

```
Error: daily_cases must have length 61
```

**Fix:** Ensure your incidence vector spans exactly days 0 through 60 (61
values). If your simulation ran for fewer days, pad with zeros or extend it.

### Packages fail to import after setup

```
[FAIL] torch: No module named 'torch'
```

**Fix:** Run `setup_python_deps(force = TRUE)` in a fresh R session. If the
problem persists, check that your internet connection is active (packages are
downloaded on first setup).

---

## Further Resources

- **Package repository:** <https://github.com/sima-njf/epiworldRcalibrate>
- **Package website:** <https://sima-njf.github.io/epiworldRcalibrate/>
- **epiworldR:** <https://github.com/UofUEpiBio/epiworldR>
- **reticulate docs:** <https://rstudio.github.io/reticulate/>
- **Reference paper:** Najafzadehkhoei, Vega Yon, Meyer and Modenesi (2025),
  [arXiv:2509.07013](https://doi.org/10.48550/arXiv.2509.07013)
