---
title: "Introduction to LSTM-based Calibration (One-Line Usage)"
author: "Sima Najafzadehkhoei"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to LSTM-based calibration}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette demonstrates the core `epiworldRcalibrate` workflow: simulate
a synthetic epidemic with known parameters, calibrate those parameters back
from the incidence curve using `calibrate_sir()`, and assess how closely the
calibrated model reproduces the original dynamics.

Using a synthetic (simulated) epidemic rather than real data is intentional —
because the ground-truth parameters are known, we can measure calibration
error directly rather than relying on goodness-of-fit alone.

::: {.callout-note}
**No Python setup required.** `calibrate_sir()` initializes the BiLSTM model
internally on first call. If you are running `epiworldRcalibrate` for the
first time on this machine, run `setup_python_deps(force=TRUE)` once beforehand.
:::

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(epiworldR)
library(epiworldRcalibrate)
```

---

## Simulate a ground-truth epidemic

We draw SIR parameters at random and simulate 60 days of epidemic dynamics.
The transmission probability `ptran` is derived from a randomly drawn `R0`
via the identity $\text{ptran} = R_0 \times \text{recov} / \text{crate}$,
ensuring internal consistency among the parameters.

The incidence vector must have **exactly 61 values** (days 0 through 60) —
this is the sequence length the BiLSTM was trained on.

```{r simulate}
set.seed(122)

n_value  <- sample(5000:10000, 1)
preval   <- runif(1, 0.007, 0.02)
crate    <- runif(1, 1, 5)
recov    <- runif(1, 0.071, 0.25)
R0_true  <- runif(1, 1.1, 5)
ptran    <- R0_true * recov / crate

true_params <- tibble(
  n                 = n_value,
  prevalence        = preval,
  contact_rate      = crate,
  transmission_rate = ptran,
  recovery_rate     = recov,
  R0                = R0_true
)

true_params
```

```{r run-true-model}
ndays      <- 60
true_model <- ModelSIRCONN(
  name              = "true_simulation",
  n                 = true_params$n,
  prevalence        = true_params$prevalence,
  contact_rate      = true_params$contact_rate,
  transmission_rate = true_params$transmission_rate,
  recovery_rate     = true_params$recovery_rate
)
run(true_model, ndays = ndays)

incidence_ts <- plot_incidence(true_model, plot = FALSE)[, 1]
stopifnot(length(incidence_ts) == 61)
```

---

## Calibrate with BiLSTM

`calibrate_sir()` passes the 61-day incidence vector through the pretrained
BiLSTM network and returns estimates for `ptran`, `crate`, and `R0` in a
single call. The recovery rate is passed in as a fixed input — consistent
with how it was fixed during simulation — so the network only needs to
recover two free parameters.

```{r calibrate}
lstm_predictions <- calibrate_sir(
  daily_cases     = incidence_ts,
  population_size = true_params$n,
  recovery_rate   = true_params$recovery_rate
)

lstm_predictions
```

We build a comparison table to see how far the estimates are from the
ground truth:

```{r params-table}
lstm_params <- tibble(
  n                 = true_params$n,
  prevalence        = true_params$prevalence,
  contact_rate      = lstm_predictions[["crate"]],
  transmission_rate = lstm_predictions[["ptran"]],
  recovery_rate     = true_params$recovery_rate,
  R0                = lstm_predictions[["R0"]]
)

params_comparison <- bind_rows(
  true_params |> mutate(param_type = "true"),
  lstm_params  |> mutate(param_type = "lstm")
)

params_comparison
```

---

## Forward simulations: true vs calibrated

To compare dynamics rather than just parameter values, we run 100 stochastic
SIR simulations under each parameter set and compute the mean trajectory with
95% prediction intervals. If the BiLSTM estimate is accurate, the two sets
of trajectories should overlap closely.

```{r forward-sims, message=FALSE}
n_reps                <- 100
all_simulation_results <- tibble()

for (i in seq_len(nrow(params_comparison))) {
  row <- params_comparison[i, ]

  fwd_model <- ModelSIRCONN(
    name              = paste0("forward_", row$param_type),
    n                 = row$n,
    prevalence        = row$prevalence,
    contact_rate      = row$contact_rate,
    transmission_rate = row$transmission_rate,
    recovery_rate     = row$recovery_rate
  )

  saver <- make_saver("total_hist")
  run_multiple(fwd_model, ndays = ndays, nsims = n_reps,
               saver = saver, nthreads = 8)
  results <- run_multiple_get_results(fwd_model, nthreads = 2)

  sim_data <- results$total_hist |>
    group_by(date, state) |>
    summarize(mean_count = mean(counts),
              ci_lower   = quantile(counts, 0.025),
              ci_upper   = quantile(counts, 0.975),
              .groups = "drop") |>
    mutate(param_type = row$param_type)

  all_simulation_results <- bind_rows(all_simulation_results, sim_data)
}
```

```{r sir-plot, fig.width=9, fig.height=10}
method_colors <- c("true" = "#440154FF", "lstm" = "#35B779FF")

sir_panel <- function(data, state_name, title) {
  pd <- filter(data, state == state_name)
  ggplot(pd, aes(x = date, color = param_type)) +
    geom_ribbon(data = filter(pd, param_type == "lstm"),
                aes(ymin = ci_lower, ymax = ci_upper, fill = param_type),
                alpha = 0.2, color = NA) +
    geom_line(aes(y = mean_count), linewidth = 1.1) +
    scale_color_manual(values = method_colors) +
    scale_fill_manual(values  = method_colors) +
    labs(title = title, x = "Day", y = "Count", color = "Method", fill = "Method") +
    theme_minimal() +
    theme(legend.position = "bottom",
          plot.title = element_text(size = 12, hjust = 0.5))
}

(sir_panel(all_simulation_results, "Susceptible", "Susceptible") /
 sir_panel(all_simulation_results, "Infected",    "Infected")    /
 sir_panel(all_simulation_results, "Recovered",   "Recovered")) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title    = "SIR Dynamics: True vs LSTM-Calibrated Parameters",
    subtitle = paste0("Mean trajectory and 95% CI across ", n_reps, " simulations")
  )
```

---

## Parameter bias

The table below shows the ground-truth values, the BiLSTM estimates, and the
absolute bias for each estimated parameter. A bias near zero means the network
recovered the true value well; larger bias values indicate where the model
struggled, which may reflect the inherent non-identifiability of the SIR
system (different combinations of `ptran` and `crate` can produce similar
incidence curves).

```{r bias-table}
tibble(
  Parameter = c("Contact Rate", "Transmission Rate", "R0"),
  True      = c(true_params$contact_rate,
                true_params$transmission_rate,
                true_params$R0),
  LSTM      = c(lstm_params$contact_rate,
                lstm_params$transmission_rate,
                lstm_params$R0)
) |>
  mutate(Bias    = round(LSTM - True, 4),
         Pct_Err = round((LSTM - True) / True * 100, 1)) |>
  knitr::kable(digits = 4,
               caption = "BiLSTM parameter estimates vs ground truth",
               col.names = c("Parameter", "True", "LSTM",
                             "Bias", "% Error"))
```
